Assessing primitives performance\\ on multi-stage execution

### 2018-05-13 paper.tex
typos

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[10pt, sigplan]{acmart}
%%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

@@

\paragraph{Baseline: pure Smalltalk.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the contructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.

@@

\begin{table*} [bth]
\centering
\begin{tabular}{c|cccc|ccc}
   &  \multicolumn{4}{c|}{Software maintenance \& evolution} & \multicolumn{3}{c}{Primitive execution time}\\
  \hline
    & Recompilation & Familiar to & Lignes & Number of & Stack switch & Optimization & Code generation \\
    & required & developers & of code & implementations & overhead & potential & control \\
  \hline
  Baseline    & No & All  & 10  & 0 & No & - & - \\
   \hline
  SmartSyntax & Plugin & VM & 20  & 3+ & Yes  & Limited & Low \\
   \hline
  SlangPlugin & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
   \hline
  PureSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
   \hline
  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
  \hline
\end{tabular}
\caption{Comparison between implementations\vspace{-0.5cm}}
\label{tab:cmp_implem}
\end{table*}

@@

\cba{In this section somewhere we need to say that criteria are split in maintenance and performance, the evaluation in next section is on last 3 criteria: Stack switch overhead, Optimization potential and control over the generated code.}

@@

\paragraph{Stack switch overhead.}
As mentionned in Section \ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure \ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performances are close to the baseline's.
As the string size goes larger, the overhead is less significant because it is **absorbed** by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time.

\paragraph{Optimization potential on large strings comparison.}
The plugins come with an overhead \sk{not an overhead when it comes to inlining. Just not optimized. need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section \ref{sec:implem_cmp}. This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}). However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster. This is not the case anymore on 1000-characters strings.
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

@@

We identified two directly related works. In some languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}. In addition, in the V8 project  \cite{V8}, the interpreter was entirely built on top of the V8 JIT's back-end using a specific DSL similar to Cog's RTL.

@@

To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition in a DSL compiled ahead of time through their JIT back-end to native code. With their primtives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives). In this context, the stack switch overhead is not a problem anymore.

>>>>

%%%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[10pt, sigplan]{acmart}
%%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan,10pt]{acmart}\settopmatter{}

@@

\paragraph{Baseline: pure Smalltalk.}
To evaluate the primitive performance, we use as baseline an implementation in pure Smalltalk. There are multiple ways of writing the comparison of two strings in Smalltalk, following Smalltalk coding conventions, a standard developer would likely use high-level constructs which apply a closure on each element of the strings. We did not use high-level construct and chose to implement the Smalltalk version the most optimized way Smalltalk allows, \emph{i.e.,} we used the constructs to:do: and ifFalse:, which are both recognized by the Smalltalk to bytecode compiler and compiled respectively to loops and branches at the bytecode level. The implementation follows the new specification, hence two methods are available, baselineCompareWith: and baselineCompareWith:order:. Since the focus is only on ASCII ordered comparison, we show the code only of baselineCompareWith:.

@@

\begin{table*} [bth]
\centering
\begin{tabular}{c|cccc|ccc}
   &  \multicolumn{4}{c|}{Software maintenance \& evolution} & \multicolumn{3}{c}{Primitive execution time}\\
  \hline
    & Recompilation & Familiar to & Lines & Number of & Stack switch & Optimization & Code generation \\
    & required & developers & of code & implementations & overhead & potential & control \\
  \hline
  Baseline    & No & All  & 10  & 0 & No & - & - \\
   \hline
  SmartSyntax & Plugin & VM & 20  & 3+ & Yes  & Limited & Low \\
   \hline
  SlangPlugin & Plugin & VM & 35  & 1  & Yes & Limited & Low \\
   \hline
  PureSlang   & Yes  & VM & 45  & 1  & Yes & High & Low \\
   \hline
  Slang+JIT   & Yes  & VM-JIT & 85  & 2  & No & High & High \\
  \hline
\end{tabular}
\caption{Comparison between implementations\vspace{-0.5cm}}
\label{tab:cmp_implem}
\end{table*}

@@

\cba{In this section somewhere we need to say that criteria are split in maintenance and performance, the evaluation in next section is on last three criteria: Stack switch overhead, Optimization potential and control over the generated code.}

@@

\paragraph{Stack switch overhead.}
As mentioned in Section~\ref{sec:VMexec}, all implementations written in Slang come with an overhead due to the switch between the C and Smalltalk stacks. This overhead is clearly visible in the bench results in Figure \ref{fig:bench} for small strings (size 1 or less): the SmartSyntax and Plugin implementations are around 30 to 60\% slower than the baseline's, and the PureSlang performances are close to the baseline's.
As the string size goes larger, the overhead is less significant because it is absorbed by other performance gains.
Regarding this criteria, the Slang+JIT version is always the fastest: it is not really impacted by the switch overhead, because the Slang part is used for a very short time.

\paragraph{Optimization potential on large strings comparison.}
The plugins come with an overhead \sk{not an overhead when it comes to inlining. Just not optimized. need rephrasing} due to the cross-file inlining and the passage through an InterpreterProxy, like said in Section~\ref{sec:implem_cmp}. This is one of the factor explaining the performance gap between these versions and the Slang and Slang+JIT ones.
When it comes to comparing large strings (100 characters and more), the execution time is mainly spent in the comparison loop (Figure \ref{fig:Asm}). However, the switch overhead is still noticeable in the PureSlang implementation when strings of 100 characters are compared: PureSlang is 15.8 times faster while Slang+JIT is 16.8 times faster. This is not the case anymore on 1000-characters strings.
This last observation only applies to the GCC results: the code generated by the LLVM compiler induces a longer execution time, as explained in the next paragraph.

@@

We identified two directly related works. In some languages (Strongtalk, Self, RSqueak), the implementors tried to decrease the overall number of optional primitives by improving the client language performance, reducing the need for such primitives \cite{Strongtalk, ursPHD, NoPrimTracing}. In addition, in the V8 project~\cite{V8}, the interpreter was entirely built on top of the V8 JIT's back-end using a specific DSL similar to Cog's RTL.

@@

To avoid paying the stack switch cost between the client Stack and the C stack, the V8 team recently wrote an interpreter Ignition in a DSL compiled ahead of time through their JIT back-end to native code. With their primitives and their interpreter written in the DSL, the V8 runtime rarely needs to switch between the client and the C stack (no need to switch to interpret code, no need to switch for primitives). In this context, the stack switch overhead is not a problem anymore.
